A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-22 doc/paper.tex
Methodology stuff. Fixes #49, fixes #50, fixes #51, fixes #52, fixes #55, fixes #56, fixes #57, fixes #58, fixes #59, fixes #60

	\begin{figure*}[t] \includegraphics[width = \textwidth]{IFargslabel} \caption{A
			picture of the tool.} \centering \label{fig:fullpic} \end{figure*}

@@

	\section{Approach/Methodology}

	\subsection{Design Goals} In visualizing the spreadsheet data, we outlined the
	core goals for what the tool should accomplish. Furthermore, with these goals,
	we also considered the likely challenges of achieving those goals and
	particular outcomes which we considered beyond the scope of the research. We
	present both sides of these goals below:

@@

	\subsection{Walkthrough of the Tool} Before discussing the minutiae of how the tool
	was developed, it will help first to give a basic example of how the tool can
	be used and how to interpret its symbols. One such example might be to find the
	answer to this question: \textit{ What kinds of functions do people use to
		define the condition in an IF function?} The final visualization is available
	in figure \ref{fig:fullpic}; however, to understand the interaction, we will
	start rather from the beginning. \par

@@

	\begin{figure} \centering \includegraphics{start} \label{fig:startpic}
		\caption{How the IF tree looks at the start} \end{figure}

	The visualization so far comprises two types of node: the circle, which
	represents a discrete function in the formula and is size according to its
	relative frequency (the top-level node, by definition, will be the largest);
	and the numbered squares, which represent the positions of arguments within its
	parent function. From this, we can infer that, of the times it was observed, IF
	can have at most three arguments passed into it, which corresponds with its
	specifications in the
	API.\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2} \par

	Knowing that it is the first argument which contains the conditionals, we click
	on the square labeled "1" to explore. To save space, when more than 10 unique
	arguments have been observed in any position, the tool displays only the first
	ten, with an option to display the rest. The results are shown below:

	\begin{figure}
		\centering
		\includegraphics[width=.43\textwidth]{IFexpand}
		\label{fig:expandif}
		\caption{First argument of the IF tree}
	\end{figure}

	As expected, we see that IF contains as its first argument a number of
	comparison operators, such as = and <=, and boolean-returning functions, like
	ISNA and AND, with simple equality being the most common and some use as ABS
	being the least seen of everything actually used. \par

	From here, we can further explore the common options among these functions.
	Clicking on the "=" node will yield two arguments, it being a binary operator,
	and expanding each of them will peer into the range of common values of
	equality comparison, which Figure \ref{fig:fullpic} shows. \par

@@

	\subsection{Design Decisions} \label{ssec:decisions}
	Early in the process, we decided that the tree
	form would be a suitable fit to represent the parent-child or caller-callee
	relationships inherent in the data, given the composition of formulae as
	functions and their arguments (which could be yet more functions). By adapting
	this structure to accommodate a broad range of possibilities for nested
	functions, the branching factor depends on both the number of arguments in a
	function and the number of possible functions observed as an argument in a
	function. \par

	Guided by the goals in section A, we faced a number of decision points, a
	sample of which we describe below, before we arrived at the design shown above.
	\begin{itemize} \item \textit{Copied formulae}: Excel allows users to spread a
		formula over an area, repeating the same task in each cell with minor
		adjustments. Without checking for this, the analysis may not reveal the
		functions most commonly used together but rather the formulae most often
		applied to large areas. To combat this, we converted formulae from their native
		A1 format to the relative R1C1, in which copied formulae should be identical,
		and reduced the records to unique R1C1 formulae per sheet. \item
		\textit{Importance of depth}: When a function appears within another, should it
		be analyzed only as a nested function or would it also be valid to analyze the
		nested function on its own?

		\par For example, in the pictured IF function, we found that people often use
		another IF statement as the second or third arguments of the top-level IF. If
		we only consider functions exactly where they're found embedded in the
		formula, then information about the same function -- IF, in this case -- will
		be scattered across different trees with no way to aggregate them. If we
		record every instance of a function by ignoring their context -- that is,
		including an IF embedded within a SUM function in the same node as the
		top-level IF -- then the tool will represent some functions in multiple nodes
		to capture every possible level of nesting. Both approaches have benefits and
		drawbacks, and so we included both. \item \textit{Pattern density}: How should
		the tool quantitatively order its elements: by a function's raw frequency or
		by the unity of patterns it leads to? For example, if SUM has for its first
		argument two possibilities, one which itself contains 1000 unique argument
		possibilities with 1 occurrence each (high frequency, low pattern density) and
		another seen with 2 argument possibilities of 100 occurrences each (low
		frequency, high pattern density), which would be more interesting to
		emphasize? The answer depends on the nuances of the questions, but for
		simplicity, I've shown the former. \item \textit{Non-functions}: How should
		the tool represent everything in the formula that isn't a function: numbers,
		string literals, errors, references, etc? Since these don't accept arguments,
		they will adorn the tree as leaves, and their precise content won't affect the
		functions around them as long as their types are known. As such, in the
		visualization, all of these nodes are replaced and aggregated under their
		types and represented as empty white nodes. \item \textit{Optional arguments}:
		How should the tool handle functions which accept a variable number of
		arguments? SUM, for example, can have anywhere from 1 to 255, and IF can
		accept either 2 or 3.
		\footnote{https://support.office.com/en-us/article/Excel-functions-by-category-5f91f4e9-7b42-46d2-9bd1-63f26a86c0eb} Without prior evidence, it's possible that, when people use optional arguments, they use different kinds of types and functions for arguments versus the cases when they ignore optional arguments. To account for this, we separate and analyze the different quantities of arguments observed for each function; the tool, however, uses as default all of the options collapsed into a single representation. \end{itemize}

	\subsection{Implementation Details} We can view the final visualization as the
	product of two discrete processes: \begin{itemize} \item \textit{Collection}:
		Given a set of Excel sheets, the tool, written mostly in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form by which it is recorded. When all formulae have
		been analyzed like this, it produces JSON files for each top-level function in
		the set. \item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily for
		its accessibility: given the JSON, the visualization can be rather simply
		embedded into a webpage accessible through a browser. \end{itemize} A
	demonstration of the tool can be found at [LINK HERE].

@@

	\begin{figure*} \centering \includegraphics[width=\textwidth]{longsum}
		\label{"fig:longsum"} \caption{The horizontal length of a tree may indicate
			complex or redundant design.} \end{figure*}

@@

	Given access to the same dataset, how might this tool approach the issue? On
	the surface, the tool supports easy quantification through the tooltips
	displayed for each node, as shown in figure \ref{"fig:vhlookups"}, answering
	the question, "How often do spreadsheet programmers use this function?". But,
	more to the point, the tool's distinction between the numbers of arguments for
	a single function lets the user compare frequencies of three-argument forms
	with the four-argument. The visualization does not, however, discern between
	the different values for booleans in the tree, much less whether the
	spreadsheet columns are ordered -- but the linking of examples to patterns
	allows quick insight into which files are at risk for the problems they
	outlined. \par

	\begin{figure} \centering \includegraphics[width=.5\textwidth]{vlookup}
		\includegraphics[width=.225\textwidth]{vlookupbox}
		\includegraphics[width=.225\textwidth]{hlookupbox}
		\includegraphics[width=.5\textwidth]{hlookup} \label{"fig:vhlookups"}
		\caption{VLOOKUP and HLOOKUP with four arguments. Numbers not expected to
			correspond perfectly with \cite{hermans2015detecting}.} \end{figure}

@@

	\begin{figure} \centering \includegraphics[width=.5\textwidth]{lookup-2}
		\includegraphics[width=.5\textwidth]{lookup-3} \label{"fig:lookups"}
		\caption{LOOKUP with two (6 occurrences) and three (1648) arguments.}
	\end{figure}

@@

	\begin{wrapfigure}{l}{.12\textwidth} \centering \includegraphics{SUM}
		\label{"fig:sum"} \end{wrapfigure}

@@

	\begin{figure} \centering \includegraphics{longIF} \label{"fig:longif"}
		\caption{The many possibilities of IF, argument 3, demonstrate the problem of
			having too many options.} \end{figure}

@@

	A larger and related problem is the loss of argument combination. For example,
	figure \ref{"fig:index"} shows the situation of the INDEX tree, wherein each
	argument position has a few distinct possibilities. While it is easy to break
	down, for each position, which argument types are most common (ranges, MATCH,
	and numbers, respectively), the tree is silent on whether MATCH functions are
	used together in the second and third arguments, among other concerns. Though
	these argument combinations could be pieced together by manually inspecting the
	lists of examples, it is nevertheless inconvenient, though not outside the
	realm of future improvement.

	\begin{figure} \centering \includegraphics[width=.5\textwidth]{index}
		\label{"fig:index"} \caption{A difficult question: how often is MATCH used in
			both the second and third arguments?} \end{figure}

>>>>

	\begin{figure*}[t] \includegraphics[width = \textwidth]{IFargslabel} \caption{A
			picture of \toolname} \centering \label{fig:fullpic} \end{figure*}

@@

	\section{Approach}

	\subsection{Design Goals} \label{goals} In visualizing the spreadsheet data, we outlined the
	core goals for what the tool should accomplish. Furthermore, with these goals,
	we also considered the likely challenges of achieving those goals and
	particular outcomes which we considered beyond the scope of the research. We
	present both sides of these goals below:

@@

	\subsection{Walkthrough of the Tool} Before discussing the minutiae of tool's development,
	though, we will give a basic example of how to use the tool. For this, we will answer this question: \textit{ What kinds of functions do people use to
		define the condition in an IF function?} Thefull  visualization that answers this 
	is available in figure \ref{fig:fullpic}; however, to understand the interaction, we will
	start from the beginning. \par

	First, a word on the IF function. In Excel, it serves as one of the logical functions: given
	a condition as the first argument that evaluates to either true or false, it takes on
	the value of either the second or third argument, respectively. Or, as the documentation
	writes out, "IF(Something is True, then do something, otherwise do something 
	else)"\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2?ui=en-US\&rs=en-US\&ad=US}.
	Furthermore, the third argument is optional; if left out, the function will return a default
	"FALSE" value if necessary. Examples of this function in practice can be seen in \ref{fig:ifexample}.

	\begin{figure}[h]
		\centering
		\includegraphics[width=.5\textwidth]{ifExample}
		\label{fig:ifexample}
		\caption{Configurations of the IF function}
	\end{figure}

@@

	\begin{figure}[h] \centering \includegraphics{start} \label{fig:startpic}
		\caption{How the IF tree looks at the start} \end{figure}

	The visualization so far comprises two types of node: the circle \includegraphics{glossary-greenonly}, which
	represents a discrete function in the formula and is size according to its
	frequency relative to the number of times to root function appeared\footnote{The root node, by definition, will
		always be the largest node in a tree.};
	and the numbered squares \includegraphics{glossary-blue}, which represent the positions of arguments within its
	parent function. From this, we can confirm that, of the times it was observed, IF
	can have at most three arguments passed into it, which corresponds with its
	specifications in the
	API.\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2} \par

	Knowing that it is the first argument which contains the conditionals, we click
	on the square labeled "1" to explore. To save space, when more than 10 unique
	arguments have been observed in any position, the tool displays only the first
	ten, with an option to display the rest as the arrow \includegraphics{glossary-arrow}.

	\begin{figure}[h]
		\centering
		\includegraphics[width=.43\textwidth]{IFexpand}
		\label{fig:expandif}
		\caption{First argument of the IF tree}
	\end{figure}

	We see that IF contains as its first argument a number of
	comparison operators, such as = and $\le$, and boolean-returning functions, like
	ISNA and AND, with simple equality being the most common and some use as ABS
	being the least seen of everything actually used. \par

	From here, we can further explore the common options among these functions.
	Clicking on the "=" node will yield two arguments
	and expanding each of them will peer into the range of common values of
	equality comparison, which Figure \ref{fig:fullpic} shows. \par

@@

	\subsection{Icon Glossary}

	From these images, we can provide a quick glossary of images within the visualization
	to aid interpretation.
	\begin{itemize}
		\item \includegraphics{glossary-green} - \textbf{Function nodes} are green
		circles labeled with the function name. Their size is determined by their
		relative frequency, scaled logarithmically, to the frequency of the root node
		in the tree. They may have any number of arguments, including zero, as
		determined by the API.

		\item \vspace{.25cm}\includegraphics{glossary-leaf} - \textbf{Value nodes} are white circles
		with a generic label. They represent elements of a formula which are which
		accept no arguments, like references (A1), ranges (A1:A10)\footnote{That is,
			only if both sides of a references are cell references. If either side is a
			function that returns a cell reference, like INDIRECT, then the colon should
			be a function node.}, numbers, boolean values, strings, and errors. Functions
		with zero arguments, like NOW, however, should still be considered function
		nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-blue} - \textbf{Argument nodes} are dark blue
		squares labeled with a number. The number is the position of an argument
		within the function in the parent node. In \ref{fig:expandif} for example,
		argument node 1 shows every function or value that was seen in the first
		argument.

		\item  \vspace{.25cm} \includegraphics{glossary-solidline} - \textbf{Solid lines} represent a
		function's ownership of its various arguments. As such, they connect the
		function nodes to argument nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-fadingline} - \textbf{Fading lines} represent
		the possibility of an argument position to have several different possible
		functions or values. As such, they connect argument nodes to function or value
		nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-arrow} - \textbf{Expansion arrows} appear
		within an argument node when there are over 10 possible functions or values in
		an argument position. By default, only the first 10 will be shown along with
		this arrow; clicking this will show the rest as well.

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} - \textbf{Optional
			argument trees} are variations of trees that show only formulae with a
		specific number of arguments, allowing the user to control for the presence of
		optional arguments. By default, trees with dark blue nodes contain every type
		of formula without regard how many arguments they have -- for example,
		single-argument ANDs could appear alongside 4-argument ANDs. See Figure
		\ref{fig:optional} to see it visualized.
	\end{itemize}

	\begin{figure}[h]
		\centering
		\includegraphics[width = .46\textwidth]{comparison}
		\includegraphics[]{comparison-ss}
		\label{fig:optional}
		\caption{The contents of each type of tree.}
	\end{figure}

	\subsection{Implementation Details} The final visualization is the
	product of two discrete processes: \begin{itemize} \item \textit{Collection}:
		Given a set of Excel sheets, the tool, written in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form. When all formulae have
		been analyzed like this, it produces JSON files for each top-level function in
		the set.

		\item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily for
		its accessibility: given the JSON, the visualization can be simply
		embedded into a webpage accessible through a browser. \end{itemize} A
	demonstration of the tool can be found at https://github.com/DeveloperLiberationFront/Excel-Function-Visualizer.

	\subsection{Design Decisions} \label{ssec:decisions}
	Early in the design, we chose the tree form for its inherent ability to
	represent the loose parent-children relationship of formulae as
	functions and their arguments, which could be yet more functions. By adapting
	this structure to accommodate a broad range of possibilities for nested
	functions, the branching factor depends alternatively on the number of arguments in a
	function and the number of possible functions observed as an argument in a
	function. \par

	Guided by the goals in section A, we faced a number of decision points, a
	sample of which we describe below, before we arrived at the design shown above.
	\begin{itemize}

%			\begin{itemize}
%				
%				\item [1] \textbf{Provide an interactive environment for free exploration.}
%				There are a lot of data in some of these collections, and the user should be
%				able to pick for themselves which to explore while ignoring the rest. Furthermore,
%				this exploration should not be designed around answering one specific question
%				but rather be open-ended.
%				
%				\item [!1] \textbf{Do not make judgments about good and bad design.} Though
%				users may certainly approach the tool with the intention of finding particular
%				formulae, \toolname will not prioritize any on metrics of good or bad
%				practice. Rather, it will present them agnostically, referring only to
%				frequency.
%				
%				\item [2] \textbf{Emphasize the quantitative patterns in formula
%					construction.} \toolname should address the questions of how often the
%				spreadsheet programmers used a certain function and where they used it. In
%				this way, it must show precise metrics, such as frequency of use and depth of
%				function nesting, of the dataset it conveys.
%				
%				\item [!2] \textbf{Do not overwhelm with numbers.} For popular functions
%				especially, we anticipated some particularly dense trees. Though
%				everything should be accessible in accordance with Goal 1, we should design in
%				a way that reduces noise.
%				
%				\item [3] \textbf{Promote a qualitative understanding of the patterns.}
%				\toolname should supplement its observations with concrete instances of
%				relevant formulae from the corpus. Ideally, it should even direct users to the
%				exact cell in the spreadsheets where the formula was used, contextualizing the
%				functions.
%				
%				\item [!3] \textbf{Do not attempt to explain formulae.} Though \toolname tries
%				to foster understanding by linking pattern to example, it should not provide
%				a precise description of what a complex formula accomplishes. The tool's user
%				must infer this on their own.
%				
%			\end{itemize}

		\item \textit{Copied formulae}: Excel allows users to spread a
		formula over an area, repeating the same task in each cell with minor
		adjustments. Without checking for this, the analysis may not reveal the
		functions most commonly used together but rather the formulae most often
		applied to large areas. To combat this, we converted formulae from their native
		A1 format to the relative R1C1, in which copied formulae should be identical,
		and considered only each R1C1 formulae once per worksheet.

		\item \textit{Importance of depth}: When a function appears within another, should it
		be analyzed only as a nested function or would it also be valid to analyze the
		nested function on its own? \par

		\ \ For example, in the pictured IF function, we found that people often use
		another IF statement as the second or third arguments of the top-level IF. If
		we only consider functions exactly where they're found embedded in the
		formula, then information about the same function -- IF, in this case -- will
		be scattered across different trees with no way to aggregate them. If we
		record every instance of a function by ignoring their context -- that is,
		including an IF embedded within a SUM function in the same node as the
		top-level IF -- then the tool will represent some functions in multiple nodes
		to capture every possible level of nesting. For now, we only include the first,
		with the latter to be added as an option in the future.

@@

	\begin{figure*}[h] \centering \includegraphics[width=\textwidth]{longsum}
		\label{"fig:longsum"} \caption{The horizontal length of a tree may indicate
			complex or redundant design.} \end{figure*}

@@

	Given access to the same dataset, how might this tool approach the issue? On
	the surface, the tool supports easy quantification through the tooltips
	displayed for each node, as shown in figure \ref{fig:vhlookups}, answering
	the question, "How often do spreadsheet programmers use this function?". But,
	more to the point, the tool's distinction between the numbers of arguments for
	a single function lets the user compare frequencies of three-argument forms
	with the four-argument. The visualization does not, however, discern between
	the different values for booleans in the tree, much less whether the
	spreadsheet columns are ordered -- but the linking of examples to patterns
	allows quick insight into which files are at risk for the problems they
	outlined. \par

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{vlookup}
		\includegraphics[width=.225\textwidth]{vlookupbox}
		\includegraphics[width=.225\textwidth]{hlookupbox}
		\includegraphics[width=.5\textwidth]{hlookup} \label{fig:vhlookups}
		\caption{VLOOKUP and HLOOKUP with four arguments. Numbers not expected to
			correspond perfectly with \cite{hermans2015detecting}.} \end{figure}

@@

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{lookup-2}
		\includegraphics[width=.5\textwidth]{lookup-3} \label{fig:lookups}
		\caption{LOOKUP with two (6 occurrences) and three (1648) arguments.}
	\end{figure}

@@

	\begin{wrapfigure}{l}{.12\textwidth} \centering \includegraphics{SUM}
		\label{fig:sum} \end{wrapfigure}

@@

	\begin{figure} \centering \includegraphics[scale=.5]{longIF} \label{fig:longif}
		\caption{The many possibilities of IF, argument 3, demonstrate the problem of
			having too many options.} \end{figure}

@@

	A larger and related problem is the loss of argument combination. For example,
	figure \ref{fig:index} shows the situation of the INDEX tree, wherein each
	argument position has a few distinct possibilities. While it is easy to break
	down, for each position, which argument types are most common (ranges, MATCH,
	and numbers, respectively), the tree is silent on whether MATCH functions are
	used together in the second and third arguments, among other concerns. Though
	these argument combinations could be pieced together by manually inspecting the
	lists of examples, it is nevertheless inconvenient, though not outside the
	realm of future improvement.

	\begin{figure} \centering \includegraphics[width=.5\textwidth]{index}
		\label{fig:index} \caption{A difficult question: how often is MATCH used in
			both the second and third arguments?} \end{figure}
